### 2016-06-21 paper/icfp16-cameraready/abstract.tex
tweaks based on wes comments

\begin{abstract}
  %
  \emph{Static} type errors are a common stumbling block
  for newcomers to typed functional languages.
  %
  We present a \emph{dynamic} approach to explaining type
  errors by generating counterexample witness inputs that
  illustrate \emph{how} an ill-typed program goes wrong.
  %
  First, given an ill-typed function, we symbolically
  execute the body to dynamically synthesize witness values
  that can make the program go wrong.
  We prove that our procedure synthesizes
  \emph{general witnesses} in that if a witness is
  found, then \emph{for all} inhabited input types, there exist
  values that can make the function go wrong.
  %
  Second, we show how to extend the above procedure to
  produce a \emph{reduction graph} that can be used to
  interactively visualize and debug witness executions.
  %
  Third, we evaluate our approach on two data sets
  comprising over 4,500 ill-typed student programs.
  Our technique is able to generate witnesses for
  88\% of the programs, and our reduction graph
  yields small counterexamples for 81\% of the witnesses.
\end{abstract}

>>>>

\begin{abstract}
  %
  \emph{Static} type errors are a common stumbling block
  for newcomers to typed functional languages.
  %
  We present a \emph{dynamic} approach to explaining type
  errors by generating counterexample witness inputs that
  illustrate \emph{how} an ill-typed program goes wrong.
  %
  First, given an ill-typed function, we symbolically
  execute the body to dynamically synthesize witness values
  that can make the program go wrong.
  We prove that our procedure synthesizes
  \emph{general witnesses} in that if a witness is
  found, then \emph{for all} inhabited input types, there exist
  values that can make the function go wrong.
  %
  Second, we show how to extend the above procedure to
  produce a \emph{reduction graph} that can be used to
  interactively visualize and debug witness executions.
  %
  Third, we evaluate the coverage of our approach on two data sets
  comprising over 4,500 ill-typed student programs.
  Our technique is able to generate witnesses for
  88\% of the programs, and our reduction graph
  yields small counterexamples for 81\% of the witnesses.
  %
  Finally, we evaluate the utility of our witnesses in
  helping students understand and fix type errors, and
  find that students presented with our witnesses
  show a greater understanding of type errors
  than those presented with a standard error message.
\end{abstract}

### 2016-06-21 paper/icfp16-cameraready/evaluation.tex
tweaks based on wes comments

\paragraph{Evaluation Goals}
%
There are three questions we seek to answer with our evaluation:
%
\begin{enumerate}
\item \emphbf{Witness Coverage}
      How many ill-typed programs can we find witnesses for?
\item \emphbf{Witness Complexity}
      How \emph{complex} are the traces produced by the witnesses?
\item \emphbf{Witness Utility}
      How \emph{helpful} are the witnesses and traces in debugging type errors?
\end{enumerate}

\paragraph{Benchmarks}
We answer the first two questions on two sets of ill-typed programs,
\ie\ programs that were rejected by the \ocaml\ compiler because of a
type error.
%
The first dataset comes from the Spring 2014 undergraduate Programming
Languages course at UC San Diego.
%
We recorded each interaction with the \ocaml\ top-level system over the
course of the first three assignments (IRB
% \# hidden for blind review),
\#140608),
from which we extracted \ucsdsize\ distinct, ill-typed \ocaml\ programs.
%
The second dataset --- widely used in the literature --- comes from a
graduate-level course at the University of Washington~\cite{Lerner2006-pj},
from which we extracted 284 ill-typed programs.
%
Both datasets contain relatively small programs, the largest being 348
SLoC; however, they demonstrate a variety of functional programming
idioms including (tail) recursive functions, higher-order functions,
polymorphic data types, and expression evaluators.

@@

\input{gallery2}

\subsection{Measuring Witness Utility}
\label{sec:user-study}
Finally, to test the explanatory power of our jump-compressed traces, we
ran a user study (IRB \#XXXX) at the University of Virginia (UVA).
%
We included four problems in an exam in the Spring session of UVA's
undergraduate Programming Languages course (CS XXX).
%
We presented the students with ill-typed \ocaml\ programs and
asked them to
%
(1) \emph{explain} the type error, and
%
(2) \emph{fix} the type error.
%
For each problem the student was given the ill-typed program and
either \ocaml's error message or \toolname's jump-compressed trace.
%
The four problems assigned to the students were the @sumList@,
@digitsOfInt@, and @wwhile@ programs from \S~\ref{sec:advantage-traces},
as well as the following @append@ program
%
\begin{ecode}
  let append x l =
    match x with
    | []   -> l
    | h::t -> h :: t :: l
\end{ecode}
%
which triggers an occurs-check error on line 4.
%
We then instructed four annotators (one of whom is an author, the other
three are teaching assistants at UCSD) to classify the answers as
correct or incorrect.

We performed an inter-rater reliability (IRR) analysis to determine the
degree to which the annotators consistently graded the exams.
%
As we had more than two annotators assigning nominal (``correct'' or
``incorrect'') ratings we used Fleiss' kappa~\cite{Fleiss1971-du} to
measure IRR.\@
%
Fleiss' kappa is measured on a scale from $1$, indicating total
agreement, to $-1$, indicating total disagreement, with $0$ indicating
random agreement.

@@

\subparagraph{Conclusion}
%
We collected exams from 60 students, though due to the nature of the
study not every student completed every problem. The number of
incomplete submissions ranges from 2 (for the \ocaml version of
@sumList@) to 17 (for the \toolname version of @wwhile@). Collecting
more responses per test pair was not possible, as it would require
having students answer the same problem twice (once with \ocaml and once
with \toolname).

\subsubsection{Results}
%
Figure~\ref{fig:results-user-study} summarizes a single annotator's
results, which show that students given \toolname's jump-compressed
trace were consistently more likely to correctly explain and fix the
type error than those given \ocaml's error message.
%
The measured kappa values were $\kappa = 0.72$ for the explanations and
$\kappa = 0.83$ for the fixes; while there is no formal notion for what
consititutes strong agreement~\cite{Krippendorff2012-wd}, kappa values
above $0.60$ are often called ``substantial''
agreement~\cite{Landis1977-ey}.
%
\begin{figure*}[ht]
\centering
\includegraphics[width=0.49\linewidth]{user-study-reason.png}
\includegraphics[width=0.49\linewidth]{user-study-fix.png}
\caption{A classification of students' explanations and fixes for type
  errors, given either \ocaml's error % message
  or \toolname's
  jump-compressed trace. The students given \toolname's jump-compressed
  trace consistently scored better than those given \ocaml's type
  error.}
\label{fig:results-user-study}
\end{figure*}
%

>>>>

\paragraph{Evaluation Goals}
%
There are three questions we seek to answer with our evaluation:
%
\begin{enumerate}
\item \emphbf{Witness Coverage}
      How many ill-typed programs can we find witnesses for?
\item \emphbf{Witness Complexity}
      How \emph{complex} are the traces produced by the witnesses?
\item \emphbf{Witness Utility}
      How \emph{helpful} (qualitatively and quantitatively)
      are the witnesses and traces in debugging type errors?
\end{enumerate}

\paragraph{Benchmarks}
We answer the first two questions on two sets of ill-typed programs,
\ie\ programs that were rejected by the \ocaml\ compiler because of a
type error.
%
The first dataset comes from the Spring 2014 undergraduate Programming
Languages (CSE 130) course at UC San Diego.
%
We recorded each interaction with the \ocaml\ top-level system over the
course of the first three assignments (IRB
% \# hidden for blind review),
\#140608),
from which we extracted \ucsdsize\ distinct, ill-typed \ocaml\ programs.
%
The second dataset --- widely used in the literature --- comes from a
graduate-level course at the University of Washington~\cite{Lerner2006-pj},
from which we extracted 284 ill-typed programs.
%
Both datasets contain relatively small programs, the largest being 348
SLoC; however, they demonstrate a variety of functional programming
idioms including (tail) recursive functions, higher-order functions,
polymorphic data types, and expression evaluators.

We answer the third question in two steps.
%
First, we present a qualitative evaluation of \toolname's traces on a
selection of programs drawn from the UCSD dataset.
%
Second, we present a quantitative user study of students in the
University of Virginia's Spring 2016 undergraduate Programming Languages
(CS 4501) course.
%
As part of an exam, we presented the students with ill-typed \ocaml\
programs and asked them to
%
(1) \emph{explain} the type error, and
%
(2) \emph{fix} the type error (IRB \#2014009900).
%
For each problem the students were given the ill-typed program and
either \ocaml's error message or \toolname's jump-compressed trace.

@@

\subsection{Witness Utility}
\input{gallery2}

\subsubsection{Quantitative}
\label{sec:user-study}
% Finally, to test the explanatory power of our jump-compressed traces, we
% ran a user study (IRB \#2014009900) at the University of Virginia (UVA).
% %
% We included four problems in an exam in the Spring session of UVA's
% undergraduate Programming Languages course (CS 4501).
% %
% We presented the 60 students in the course with ill-typed \ocaml\
% programs and asked them to
% %
% (1) \emph{explain} the type error, and
% %
% (2) \emph{fix} the type error.
% %
% For each problem the student was given the ill-typed program and
% either \ocaml's error message or \toolname's jump-compressed trace.
%
We assigned four problems to the ($n=60$) students in the course: the
@sumList@, @digitsOfInt@, and @wwhile@ programs from
\S~\ref{sec:advantage-traces}, as well as the following @append@ program
%
\begin{ecode}
  let append x l =
    match x with
    | []   -> l
    | h::t -> h :: t :: l
\end{ecode}
%
which triggers an occurs-check error on line 4.
%
For each problem the students were given the ill-typed program and
either \ocaml's error message or \toolname's jump-compressed trace.
%
Due to the nature of an in-class exam, not every student answered every
question; we received between 13 and 28 (out of a possible 30) responses
for each problem-tool pair.

We then instructed four annotators (one of whom is an author, the other
three are teaching assistants at UCSD) to classify the answers as
correct or incorrect.
%
We performed an inter-rater reliability (IRR) analysis to determine the
degree to which the annotators consistently graded the exams.
%
As we had more than two annotators assigning nominal (``correct'' or
``incorrect'') ratings we used Fleiss' kappa~\cite{Fleiss1971-du} to
measure IRR.\@
%
Fleiss' kappa is measured on a scale from $1$, indicating total
agreement, to $-1$, indicating total disagreement, with $0$ indicating
random agreement.

@@

\subparagraph{Conclusion}
%
We collected exams from 60 students, though due to the nature of the
study not every student completed every problem.
%
The number of complete submissions ranges from 13 (for the \toolname
version of @wwhile@) to 28 (for the \ocaml version of @sumList@), out of
a maximum of 30 per program-tool pair.
%
Collecting more responses per test pair was not possible, as it
would require having students answer the same problem twice (once with
\ocaml and once with \toolname).

\paragraph{Results}
%
Figure~\ref{fig:results-user-study} summarizes a single annotator's
results, which show that students given \toolname's jump-compressed
trace were consistently more likely to correctly explain
and fix the type error than those given \ocaml's error message.
%
Across each problem the \toolname responses were marked correct
$10-30\%$ more often than the \ocaml responses, which suggests that
the students who had access to \toolname's traces had a better
understanding of the type errors.
%
The measured kappa values were $\kappa = 0.72$ for the explanations and
$\kappa = 0.83$ for the fixes; while there is no formal notion for what
consititutes strong agreement~\cite{Krippendorff2012-wd}, kappa values
above $0.60$ are often called ``substantial''
agreement~\cite{Landis1977-ey}.
%
\begin{figure*}[ht]
\centering
\includegraphics[width=0.49\linewidth]{user-study-reason.png}
\includegraphics[width=0.49\linewidth]{user-study-fix.png}
\caption{A classification of students' explanations and fixes for type
  errors, given either \ocaml's error % message
  or \toolname's
  jump-compressed trace. The students given \toolname's jump-compressed
  trace consistently scored better ($\ge 10\%$) than those given \ocaml's type
  error.}
\label{fig:results-user-study}
\end{figure*}
%
